Scaling rates of true polar wander in convecting planets and moons

### 2016-06-06 tpw_rate.tex
Fix statement on choice of coordinate system.

The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_2$ plane ($\phi = 0$).
Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
\begin{equation}
\vert \dot{\Theta} \vert = 
\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
\label{eq:simple_milankovitch}
\end{equation}
The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).

@@

Two processes control the evolution of $\theta$. Its growth occurs through perturbations in the convective moment and 
its decay occurs by relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space in Equation~\eqref{eq:euler}).
The $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, 
which we describe by the rotation vector $\mathbf{\Psi}$. 
(To be explicit in our discussion we let $\mitbf{\Psi}$ define the rotation of the $\mathbf{E}$-frame in the geographic frame, 
although this choice is arbitrary as long as we are consistent). The time derivatives of $\mitbf{\omega}$ in the two frames are related by 
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
where primes are used to define quantities in the $\mathbf{E}$-frame. Rearranging gives 
\begin{equation}
\dot{\mitbf{\omega}}^{\prime} = \dot{\mitbf{\omega}} - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
We now substitute for $\dot{\mitbf{\omega}}$ from Equation~\eqref{eq:liouville_ode}, 
making the additional assumption that the geographic and $\mathbf{E}$-frames are momentarily aligned. 
This assumption is not a serious restriction because our goal is simply to relate the time derivatives of $\mitbf{\omega}$ in the two frames, 
rather than track the evolution of $\mitbf{\omega}$ over time. The expression for $\dot{\mitbf{\omega}}$ becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing the rate of TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we can choose a convenient coordinate system by aligning the axes so that 
$\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
In this case the orientation of $\mitbf{\omega}$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
The orientation of $\Psi$ is not restricted in this frame, 
so we specify the direction using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).

>>>>

The second special case is if the rotation vector $\mitbf{\omega}$ lies in $\mathbf{e}_1$-$\mathbf{e}_3$ plane ($\phi = 0$).
Then Equation~\eqref{eq:polar_wander_rate} becomes significantly simpler, which is useful for scaling purposes:
\begin{equation}
\vert \dot{\Theta} \vert = 
\frac{1}{2(C-A) T_1}\sin{2\theta} ( \lambda_3 - \lambda_1).
\label{eq:simple_milankovitch}
\end{equation}
The maximum polar wander rate of the system is achieved when $\theta=45^\circ$ (see, e.g. \citet{fung1965foundations}).

@@

Two processes control the evolution of $\theta$. Its growth occurs through perturbations in the convective moment and 
its decay occurs by relaxation of the pole towards the maximum moment of inertia.
We can explore these two effects by converting Equation~\eqref{eq:liouville_ode}
from the body-fixed frame to the $\mathbf{E}$-frame (recall that $\mitbf{\Omega}$ defines
the rotation of the body-fixed frame relative to inertial space in Equation~\eqref{eq:euler}).
The $\mathbf{E}$-frame rotates slowly with respect to the geographic frame, 
which we describe by the rotation vector $\mathbf{\Psi}$. 
(To be explicit in our discussion we let $\mitbf{\Psi}$ define the rotation of the $\mathbf{E}$-frame in the geographic frame, 
although this choice is arbitrary as long as we are consistent). The time derivatives of $\mitbf{\omega}$ in the two frames are related by 
\begin{equation}
\dot{\mitbf{\omega}} = \dot{\mitbf{\omega}}^\prime + \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
where primes are used to define quantities in the $\mathbf{E}$-frame. Rearranging gives 
\begin{equation}
\dot{\mitbf{\omega}}^{\prime} = \dot{\mitbf{\omega}} - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\end{equation}
We now substitute for $\dot{\mitbf{\omega}}$ from Equation~\eqref{eq:liouville_ode}, 
making the additional assumption that the geographic and $\mathbf{E}$-frames are momentarily aligned. 
This assumption is not a serious restriction because our goal is simply to relate the time derivatives of $\mitbf{\omega}$ in the two frames, 
rather than track the evolution of $\mitbf{\omega}$ over time. The expression for $\dot{\mitbf{\omega}}$ becomes
\begin{equation}
 \dot{\mitbf{\omega}}^\prime  = \frac{1}{(C-A)T_1} \left[ \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime - \left( \mitbf{\omega}^\prime \cdot \mathbf{E}^\prime \cdot \mitbf{\omega}^\prime  \right) \mitbf{\omega}^\prime \right] - \mathbf{\Psi}\times \mitbf{\omega}^\prime.
\label{eq:liouville_ode_new_frame}
\end{equation}
The advantage of expressing the rate of TPW in the $\mathbf{E}$-frame is that the entire equation
may be written in terms of the principal moments $\lambda_i$ and the angles from the principal axes.
As in Equation~\eqref{eq:simple_milankovitch}, we make the simplifying assumption that 
$\mitbf{\omega}^{\prime}$ lies in $\mathbf{e}_1^{\prime}-\mathbf{e}_3^{\prime} $ plane (as shown in Figure~\ref{fig:reference_frames}). 
In this case the orientation of $\mitbf{\omega}$ is defined by colatitude $\theta$ and longitude $\phi=0$. 
The colatitude $\theta$ is precisely the angular misalignment that specifies the rate of TPW in Equation~\eqref{eq:simple_milankovitch}.
We furthermore specify the orientation of $\Psi$
using colatitude $\beta$ and longitude $\gamma$ (see Figure~\ref{fig:reference_frames}).
